---
title: "BCB546 R Assignment"
author: "Brittany Cavazos"
date: "September 30, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data Inspection

First I need to read in the data and briefly look over the structure of the datasets
```{r reading in data}
genotypes_raw<-read.delim("fang_et_al_genotypes.txt", header = T, sep = "\t")
dim(genotypes_raw)
head(genotypes_raw[,1:5]) # just showing the output of first five columns for all of these inspections because there are a ton of columns..
tail(genotypes_raw[,1:5])
# str(genotypes_raw) output is pretty long
snp_pos<-read.delim("snp_position.txt", header = T, sep = "\t")
dim(snp_pos)
head(snp_pos[,1:5])
tail(snp_pos[,1:5])
# str(snp_pos)
```

Next I need to transpose the data and subset to maize and teosinte datasets. 
For maize: Group = ZMMIL, ZMMLR, and ZMMMR 
```{r}
summary(genotypes_raw$Group)
  # for ZMMIL, ZMMLR and ZMMMR, we want a total maize subset that is 1573 rows long
maize_raw<-dplyr::filter(genotypes_raw, Group == "ZMMIL" | Group == "ZMMLR" | Group == "ZMMMR")
  # there is probaby a shorter way of doing this but it works

  # not to transpose it to have SNP ID as columns
names(maize_raw)[1:25]
maize_T<-as.data.frame(t(maize_raw))
dim(maize_T)
Maize_IDs<-as.character(maize_raw$Sample_ID)
colnames(maize_T)<-Maize_IDs
```

For teosinte: Group = ZMPBA, ZMPIL, and ZMPJA 
```{r}
teosinte_raw<-dplyr::filter(genotypes_raw, Group == "ZMPBA" | Group == "ZMPIL" | Group == "ZMPJA")
  # there is probaby a shorter way of doing this but it works
  # should be 975 rows after looking at summary(genotypes$Group)
dim(teosinte_raw)
  # not to transpose it to have SNP ID as columns
names(teosinte_raw)[1:25]
teosinte_T<-as.data.frame(t(teosinte_raw))
dim(teosinte_T)
Teosinte_IDs<-as.character(teosinte_raw$Sample_ID)
colnames(teosinte_T)<-Teosinte_IDs # after transposing we got weird column names so this is a way to fix that

```
Now I need to join together the snp_pos with the other two data sets, but I only need a couple columns from the snp data set
```{r}
names(snp_pos)
snp_pos<-dplyr::select(snp_pos, SNP_ID, Chromosome, Position)
  # usally writing over data frames isn't great but this is probably fine

```
Now I need to prep the data sets for merges - we will join maize/teosinte with snp_pos by snp id
```{r}

# need to add a column in both maize and teosinte that is the snp id - currently its only as the rowname
row.names(snp_pos)<-as.character(snp_pos$SNP_ID)

maize_T_forjoin<-maize_T[4:986,] # to join the snpid info we have to get rid of some of the sample info, we can tack this back on at the end if we need to
teosinte_T_forjoin<-teosinte_T[4:986,]

maizejoined<-merge(snp_pos, maize_T_forjoin, by = "row.names", all = F)
dim(maizejoined)
maizejoinedF<-dplyr::select(maizejoined, -Row.names)
teosintejoined<-merge(snp_pos, teosinte_T_forjoin, by = "row.names", all = F)
dim(teosintejoined)
teosintejoinedF<-dplyr::select(teosintejoined, -Row.names) # F for final

```

## Data Processing
_For Maize_
Now we need to make 10 files (1 for each chromosome) with SNPs ordered based on increasing position values and with missing data encoded by this symbol: ? 
# 
```{r}
head(maizejoinedF[maizejoinedF$Chromosome == 1,]) # they are currently out of order

# first let's make chromosome and position numeric
maizejoinedF$Chromosome<-as.numeric(maizejoinedF$Chromosome)
maizejoinedF$Position<-as.numeric(maizejoinedF$Position)
# sort by chromosome, then position
library(dplyr) # not totally sure if this is needed or not
maizejoined_sorted <- maizejoinedF %>% group_by(Chromosome)
maizejoined_sorted <- maizejoined_sorted %>% arrange(Position, .by_group = T)

# now subset and save data frames by chromosome (? already show missing data)

for (i in 1:10) {
  Chr_i<- dplyr::filter(maizejoined_sorted, Chromosome == i)
  filename <- paste("MaizeChromosome", i, ".txt", sep="")
  write.table(Chr_i, file = filename, sep = "\t")
}

# now reverse the sorting and replace ? with -
maizejoined_sortedR <- maizejoinedF %>% group_by(Chromosome)
maizejoined_sortedR <- maizejoined_sortedR %>% arrange(desc(Position), .by_group = T)
head(maizejoined_sortedR)

# need to make snps characters
# and then use gsub
```
